### 2018-03-02 submission/manuscript.Rmd
discussion and journal csl update

---
title: ''
csl: microbiome.csl
fontsize: 12pt
output:
  pdf_document:
    includes:
      in_header: header.tex
    keep_tex: yes
  word_document: default
geometry: margin=1.0in
bibliography: references.bib
---

@@

Select subset community you can do as well as OTUs or genera when using all of the information and that these models perform well when trained on one cohort and tested on another.

Why didn’t the whole community work better? Communities are patchy and higher level taxonomic information pools some of that patchiness
There isn’t one single bug associated with disease – why ? possibly because multiple bugs can carry out the same function (e.g. inflammation). Tell some examples (Esch Fuso, pepto)
Also, interesting that the loss of bacteria can be associated with disease. These appear to be commonly thought as beneficial bacteria that ferment to produce SCFAs

Adenomas suck. 
The associations between the microbiota and individuals with adenoma tumors are inconclusive, in part, because many studies may not be powered effectively to observe small effect sizes. None of the studies analyzed were properly powered to detect a 10% or lower change between cases and controls. The results within our meta-analysis suggest that a small effect size may well be the scope in which differences consistently occur between controls and those with adenomas. Future studies investigating adenoma tumors and the microbiota need to take power into consideration to reproducibly study whether the microbiota contributes to adenoma formation. In contrast to adenoma stage of disease, our observations suggest that most studies analyzed have sufficient power to detect many changes in the carcinoma-associated microbiota because of large effect size differences between cases and controls [Figure 5].

Which is kind of at odds with Baxter, but Baxter looked at lesions and also included FIT
Also, variation in how adenomas are defined – small polyps vs larger neopplasias (e.g. baxter), but we couldn’t do this for the current study because we didn’t have that level of detail and most studies are small
This isn’t such a novel result (Nich Chia)
Stool may not be the best place to look for early stage adenomas

Although stool represents an easy and less invasive way to assess risk, it is not clear how well this sample reflects adenoma- and carcinoma- associated microbial communities. Some studies have tried to assess this in health and disease but are limited by their sample size [@flemer_tumour-associated_2017; @Flynn_preprint_2017]. Sampling the microbiota directly associated with colon tissue may provide clearer answers but the colon tissue-based studies did not provide a clearer understanding of how the microbiota may be associated with tumors. Generally, the full OTU-based models of unmatched and matched colon tissue samples were concordant with stool samples showing that GI resident microbes were the most prevalent in the top 10 most important variables across study [Figure S3]. *Fusobacterium* was not identified consistently across studies and this could be due to both a small number of studies and a small sample size within these studies. Additionally, the majority of the colon tissue-based results were study specific with many of the top 10 taxa being present only in a single study. The presence of genera associated with contamination [@Salter_contamination_2014], within the top 10 most important variables for the genera and OTU models is worrying (i.e. *Novosphingobium*, *Acidobacteria Gp2*, *Sphingomonas*, etc.). The low bacterial biomass of tissue samples coupled with potential contamination and small sample sizes could explain why these results seem to be more sporadic than the stool results.

@@

Looking forward
Potential for using microbiome as a diagnostic tool
We’re excited that we were able to validate a set of biomarkes across multiple studies
Found a phenotype we can relate to the microbiome, it’s replicable and it’s strong

Our study identifies clear but small differences in diversity at the community level and larger differences for individual genera, present in patients with tumors versus controls [Figure 1-3]. Although there was a step-wise decrease in diversity as disease progressed from control to adenoma to carcinoma, this did not translate into large effect size increases in OR for either adenoma or carcinoma tumors. Even though mouth-associated genera increased individaul's OR of having a carcinoma for certain sample types, they did not consistently increase the OR of having an adenoma. By using these taxa that had significant ORs after multiple comparison correction we found that we could classify indviduals with either adenoma or carcinoma as well as models that use either all genera or all OTUs. Finally, many studies were individually under powered to be able to reject the null hypothesis and this could one reason only the comparison between control and carcinoma individuals for stool samples had relible detectable differences.

The data presented herein support the importance of specific taxa for carcinoma, but not necessarily adenoma, tumor formation. The results that we have presented show that the significant OR taxa model and both the full genera and OTU models, for indviduals with carcinoma, had similar AUCs [Figure 2 & 3]. This suggests that an interplay between a select number of potentially protective and exacerbating microbes within the GI community could be crucial for carcinoma formation. Importantly, it suggests that there may be key members of the GI community that should be studied further to potentially help reduce the risk of carcinoma tumor formation. Conversely, using the present data, it is clear that new approaches may be needed to identify members of the community associated with adenoma tumors. Regardless of sample type and whether a full genera- or OTU-based model was used, our Random Forest models consistently performed poorly. Yet, the step-wise decrease in diversity suggests that the adenoma-associated community is not normal but has changed subtly. This change in diversity, at this early stage of disease, could be focal to the adenoma itself. How the host interacts with these subtle changes at early stages of the disease could be what leads to a thoroughly dysfunctional community that is supportive of tumorgenesis.

For the full genera- and OTU-based models within stool, common GI microbes were most consistently present in the top 10 genera or OTUs across studies [Figure 4]. Changes in *Bacteroides*, *Ruminococcaceae*, *Ruminococcus*, and *Roseburia* were consistently found to be in the top 10 most important variables across the different studies for both individuals with adenoma and carcinoma [Figure 4]. These data suggest that whether the non-resident bacterium is *Fusobacteria* or *Peptostreptococcus* may not be as important as how these bacteria interact with the changing resident community. Based on these observations, it is possible to hypothesize that small changes in community structure lead to new niches in which any one of the mouth-associated or general inflammatory genera can gain a foothold, exacerbating the initial changes in community and facilitating the transition from adenoma to carcinoma stage of disease.

## Conclusion

By aggregating together a large collection of studies analyzing both fecal and colon tissue samples, we are able to provide evidence supporting the importance of the bacterial community in colorectal tumors. The data presented here suggests that mouth-associated microbes can gain a foothold within the colon and are commonly associated with the greatest OR of individuals having a carcinoma. Conversely, no conclusive signal with these mouth-associated microbes could be detected for individuals with an adenoma. Our observations also highlight the importance of power and sample number considerations when investigating the microbiota and adenoma tumors due to possible subtle changes in the community. Overall, the associations between the microbiota and individuals with carcinomas were much stronger than with those with adenomas.

>>>>

---
title: ''
csl: mbio.csl
fontsize: 12pt
output:
  pdf_document:
    includes:
      in_header: header.tex
    keep_tex: yes
  word_document: default
geometry: margin=1.0in
bibliography: references.bib
---

@@

Although we expected that the full OTU models would perform the best at classifying individuals with and without carcinomas, our observations suggest that both the full and significant OR taxa models performed equally well. These results suggest that lower level classification to species and strain may not add extra useful information with respect to prediction models. This has been suggested in previous literature where metagenomics did not perform better than 16S rRNA gene sequencing data at classifying individuals with normal colons and those with carcinomas [@Hannigan2017]. One possible reason as to why lower level classification may not result in better models is that the communities are patchy and higher level taxonomic information pools some of this patchiness allowing for better prediction models. There may also be a fair bit of data redundancy within models that utilize more of the community. An example of this redundancy would be when we trained models on one study and tested it on the other studies and the AUCs of the models created with the select OR taxa performed as well as full taxa models [Figure 6B].

Our observations also suggest that a small collection of taxa can classify diease as well as full OTU-based models but that these taxa individually perform quite poorly [Figure 3]. This result supports the contention that there might be redundancy of function even amongst the taxa included in the significant OR models. As an example multiple different microbes could be similarily stimulating the activation of inflammatory pathways and by doing so exacerbate disease progression. Multiple reports within the literature have found that different bacteria, such as *Escherichia coli* and *Fusobacterium nucleatum*, can worsen similarily worsen inflammation in mouse models of tumorigenesis [@ecoli_Arthur_2014; @arthur_intestinal_2012; @kostic_fusobacterium_2013]. Although the inflammatory taxa were patchy in their importance and presence across studies those that were not typically associated with inflammation were consistently important for every study [Figure 6]. The loss of these taxa (*Ruminococcus* and *Clostridium XI* in stool and *Dorea* and *Blautia* in unmatched tissue) is particularly interesting because many are commonly thought to be beneficial due to their involvement in production of short chain fatty acids [@Venkataraman2016; @Herrmann2018; @Reichardt2017].

The adenoma models as a whole performed poorly in classifying individuals with and without adenomas. This outcome is not inconsistent with what has been published previously [@hale_shifts_2017; @Sze2017]. However, the modeling results are at odds with results obtained in Baxter, *et al.* [@baxter_microbiota-based_2016]. There are some major differences between the models generated in this meta-analysis and what was used in this previous report. First, the prevoius report's models investigated the classification of lesions (individuals with adenoma or carcinoma) and not adenoma alone. The Baxter, *et al.* models also contained Fecal Immunoglobulin Test data while our meta-analysis models only contained 16S rRNA gene sequencing data. Although being able to classify individuals with adenomas is important, the classification of advanced adenomas is a more clinically meaningful diagnostic tool (i.e. those that are at high risk of progressing to a carcinoma). It is possible that we might have been able to detect differences in the bacterial community if advanced adenomas were separated from adenomas but that data was not available for the majority of studies analyzed. It is also possible that the initial changes to the bacterial community are focal to where the initial adenoma develops and would not be easily assessed with a fecal sample.

Although stool represents an easy and less invasive way to assess risk, it is not clear how well this sample reflects adenoma- and carcinoma- associated microbial communities. Some studies have tried to assess this in health and disease but are limited by their sample size [@flemer_tumour-associated_2017; @Flynn_preprint_2017]. Sampling the microbiota directly associated with colon tissue may provide clearer answers but the colon tissue-based studies did not provide a clearer understanding of how the microbiota may be associated with tumors. Generally, the full OTU-based models of unmatched and matched colon tissue samples were concordant with stool samples showing that GI resident microbes were the most prevalent in the top 10 most important variables across study [Figure S3]. *Fusobacterium* was not identified consistently across studies and this could be due to both a small number of studies and a small sample size within these studies. Additionally, the majority of the colon tissue-based results were study specific with many of the top 10 taxa being present only in a single study. The presence of genera associated with contamination [@Salter_contamination_2014], within the top 10 most important variables for the genera and OTU models is worrying (e.g. *Novosphingobium*, *Acidobacteria Gp2*, *Sphingomonas*, etc.). The low bacterial biomass of tissue samples coupled with potential contamination and small sample sizes could explain why these results seem to be more sporadic than the stool results.

@@

By aggregating together a large collection of studies analyzing both fecal and colon tissue samples, we are able to provide evidence supporting the importance of the bacterial community in carcinoma tumors. Although further validation of the biomarkers presented here need to be undertaken, the replciability of the AUC of a specific collection of taxa across multiple studies suggests a strong potential for the use of the microbiota as a risk stratification tool for individuals with carcinomas.
